pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-05-12 bmc_article.tex

Here, we develop a flexible R framework for pipeline comparison and apply it to the evalution of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Supplementary Figure 1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

We next investigated the impact of filtering according to several methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (i.e. sorted datasets such as the Zheng datasets). Indeed, the datasets where doublet removal did not have an impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that the same datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were most challenging across for all methods. An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact in the Smart-seq datasets (Koh and Kumar). When performed with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datsets, possibly owing to its regularization.  In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by the variance their explain tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).

@@

We selected the most important alternatives for each parameter (see methods) in order to explore their interactions. Analysis of variance (after correcting for baseline differences between datasets and the resolution parameter) attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality, followed by normalization, and very little to filtering (Supplementary Figure 28). Because the choice of dimensions impacts the number of clusters, we repeated this analysis of variance among clusterings that yielded the true number of clusters, and observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.

We next investigated whether the benefits of some methods where conditional on choices at other steps. Although analysis of variance identified many weak but significant interactions, in general we found that the benefits brought by important methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30). For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again indepdendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29). Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives, and the FDR nevertheless remained acceptable with 1-3 variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

Among all pipelines that were tested, the performance varied greatly among datasets. The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested. The average silhouette width of half of their subpopulations was systematically lower that the other subpopulations/ datasets and the ARI score at true number of clusters, if ever achieved, was always poor. It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1). The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}. In contrast, the Koh dataset harbors as many populations, but originates from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulations complexity and most methods applied on them performed well.

@@

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0). The glmPCA was performed using the \texti{glmpca} package (The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

@@

The `pca' and `pca2' clusters refer to the multivariate outlier detection methods implemented in \textit{scater}, running \textit{runPCA} with \texttt{use\_coldata=TRUE, detect\_outliers=TRUE}. `pca` uses all covariates, while `pca2` uses only the log10(counts), log10(features), proportion mitochondrial and proportion in the top 50 features.

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
    \fi
    \caption{\textbf{Evaluation of normalization procedures. A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. Grey squares indicate that the true number of clusters could not be reached with the corresponding method and dataset. `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping. Other evaluation metric heatmaps (e.g. panel \textbf{C} here) map colors linearly to the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median. Using this mapping, the same \textit{difference} in color has the same meaning across datasets, and the color-scale is not primarily capturing baseline differences between datasets. The numbers printed in the cells represent the raw (i.e. unscaled) metric values.}
    \label{fig:norm}
\end{figure}

>>>>

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Supplementary Figure 1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score across subpopulations. A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datasets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). PCA-based selection did not appear superior to feature-wise filters. We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering tended improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g. FACS-sorted datasets). Indeed, the datasets where doublet removal did not have a clear positive impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact among the Smart-seq datasets (Koh and Kumar). When performed with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datasets, possibly owing to its regularization.  In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).

@@

We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions. Analysis of variance (after correcting for baseline differences between datasets and the resolution parameter) attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality, followed by normalization, and very little to filtering (Supplementary Figure 28). Because the choice of dimensions impacts the number of clusters, we repeated the analysis of variance among the combinations of methods (up to clustering) that yielded the true number of clusters. We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.

We next investigated whether the benefits of some methods where conditional on the choices at other steps. Although analysis of variance identified many weak but significant interactions, in general we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30). For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29). Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with 1-3 variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions - see for instance our 'default' filters in the methods) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

Among all pipelines that were tested, the performance varied greatly among datasets. The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested. The average silhouette width of half of their subpopulations was systematically lower that the other subpopulations/ datasets and the ARI score at true number of clusters, if ever achieved, was always poor. It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1). The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}. In contrast, while the Koh dataset harbors as many populations, they originate from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulations complexity and most methods applied on them performed well.

@@

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0). The glmPCA was performed using the \texti{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

@@

The `pca.all' and `pca.sel' clusters refer to the multivariate outlier detection methods implemented in \textit{scater}, running \textit{runPCA} with \texttt{use\_coldata=TRUE, detect\_outliers=TRUE}. `pca.all` uses all covariates, while `pca.sel` uses only the log10(counts), log10(features), proportion mitochondrial and proportion in the top 50 features.

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
    \fi
    \caption{\textbf{Evaluation of normalization procedures. A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. Grey squares indicate that the true number of clusters could not be reached with the corresponding method and dataset. `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping. Other evaluation metric heatmaps (e.g. panel \textbf{C} here) map colors to the signed square-root of the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median. Using this mapping, \textit{differences} in color have the same meaning across datasets, and the color-scale is not primarily capturing outliers or baseline differences between datasets. The numbers printed in the cells represent the raw (i.e. unscaled) metric values.}
    \label{fig:norm}
\end{figure}
